---
title: "Class19: Genome Informatics"
author: 'Monica Lin (PID: A15524235)'
date: "11/30/2021"
output: pdf_document
---

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. We are interested in assessing genetic differences on a population scale. 


```{r}
expr <- read.table("Expression genotype results.txt")
head(expr)
```

> **Q13**. How many samples do we have?

```{r}
nrow(expr)
```

> What are the sample sizes for each genotype?

```{r}
table(expr$geno)
```

> What are the corresponding median expression levels for each genotype?

```{r}
boxplot <- boxplot(exp~geno, data=expr)


# The third row displays the median values for each genotype. The second value 
# designates to how many decimal places the statistics will round.

round(boxplot$stats, 2)
```

The sample size of genotype A|A is 108, with a median value of 31.25. A|G has a sample size of 233 and a median value of 25.06. G|G has a sample size of 121 and a median value of 20.07.


> **Q14**. Let's make a boxplot, with a box per genotype.

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(x=geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE) + xlab("Genotype") + ylab("Expression")
```

> What could you infer from the relative expression value between A|A and G|G displayed in this plot? Does the SNP affect the expression of ORMDL3?

The homozygous G|G genotype is down-expressed in relation to the A|A and A|G genotypes. In other words, the A|A genotype is upregulated while G|G is downregulated. The SNP lowers the expression of ORMDL3.









