---
title: "Class10_Halloween_Candy"
author: 'Monica Lin (PID: A15524235)'
date: "10/28/2021"
output: pdf_document
---

## 1. Importing candy data

```{r}
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv"
candy <- read.csv(url, row.names=1)
head(candy,n=5)
```
> **Q1**. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> **Q2**. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## 2. What is your favorite candy?

`winpercent` tells the percentage of people who prefer a given candy over another randonly chosen candy from the dataset. Higher values indicate a more popular candy.
Example code for Twix:

```{r}
candy["Twix", ]$winpercent
```

> **Q3**. What is your favorite candy in the dataset and what is its `winpercent` value?

```{r}
candy
candy["Peanut M&Ms", ]$winpercent
```

> **Q4**. What is the `winpercent` value for "Kit Kat"?

```{r}
candy["Kit Kat", ]$winpercent
```

> **Q5**. What is the `winpercent` value for "Tootsie Roll Snack Bars"?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

Use the `skim()` function to give a quick overview of a given dataset.

```{r}
install.packages("skimr")
library("skimr")
skim(candy)
```

> **Q6**. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The percent columns, particularly `winpercent`, look to be on a different scale (0-100, continuous) than the other category columns (0-1, discrete).

> **Q7**. What do you think a zero and one represent for the candy$chocolate column?

0 is no chocolate, and 1 is chocolate present in the candy.

> **Q8**. Plot a histogram of `winpercent` values

```{r}
hist(candy$winpercent)
```


> **Q9**. Is the distribution of winpercent values symmetrical?

No, the distribution is skewed right.

> **Q10**. Is the center of the distribution above or below 50%?

The center is below 50%.

> **Q11**. On average is chocolate candy higher or lower ranked than fruit candy?

First need to find all the chocolate candy rows in the `candy` dataset.

```{r}
inds <- as.logical(candy$chocolate)
chocolate <-  candy[inds,]$winpercent
mean(chocolate)
```

```{r}
inds2 <- as.logical(candy$fruity)
fruity <- candy[inds2,]$winpercent
mean(fruity)
```
Chocolate candy is ranked higher than fruit candy.

> **Q12**. Is this difference statistically significant?

```{r}
t.test(chocolate, fruity)
```
The p-value is below 0.05 threshold, signifying the difference is statistically significant.

## 3. Overall Candy Rankings

> **Q13**. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

> **Q14**. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

> **Q15**. Make a first barplot of candy ranking based on `winpercent` values.

```{r}
library(ggplot2)

ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col()
```

> **Q16**. This is quite ugly. Use the `reorder()` function to get the bars sorted by `winpercent`?

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy), winpercent)) + geom_col()
```

# Time to add some useful color

Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), orange (for candy bars) and pink (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "orange"
my_cols[as.logical(candy$fruity)] = "pink"
```

Now try the barplot with these colors.
```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```
We use the command `fill=my_cols` for `geom_col()` instead of `col=my_cols`so that there are no black outlines of each column that can confuse us with the legend.

> **Q17**. What is the worst ranked chocolate candy?

Nik L Nip

> **Q18**. What is the best ranked fruity candy?

Reeses Peanut Butter cup

# Looking at pricepercent
What about value for money? WHat is the best candy for the least money?
The regular `geom_label()` that comes with ggplot2 will make our many labels overlap and difficult to read. Instead, we can use `geom_text_repel()` function from the **ggrepel** package.

```{r}
install.packages("ggrepel")
#library("ggrepel")

# Plot price vs win
ggplot(candy) + aes(winpercent, pricepercent, label=rownames(candy)) + geom_point(col=my_cols) + geom_text_repel(col=my_cols, size=3.3, max.overlaps=5)
```

> **Q19**. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures

> **Q20**. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing=TRUE)
head(candy[ord, c(11,12)], n=5)
```

## 5. Exploring the correlation structure
Use the **corrplot** package to plot a correlation matrix

```{r}
install.packages("corrplot")
library("corrplot")
```
```{r}
cij <- cor(candy)
corrplot(cij)
```

> **Q22**. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate x Fruit are anti-correlated.

> **Q23**. Similarly, what two variables are most positively correlated?

Chocolate x Bar are most positively correlated.

## 6. Principal Component Analysis

Apply PCA using the `prcomp()` function to our candy dataset, remembering to set the `scale=TRUE` argument.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

Now plot our main PCA score plot of PC1 vs PC2.
```{r}
plot(pca$x[,1:2], col=my_cols)
```

Let's use ggplot to make a nicer version of this. Here we need a new data.frame that contails all the info we want displayed in teh plot, so we want to combine our PCA results with all the candy data.
```{r}
my_data <- cbind(candy,pca$x[,1:3])

ggplot(my_data) + aes(PC1, PC2, label=rownames(candy)) + geom_point(col=my_cols) + geom_text_repel(col=my_cols, size=3, max.overlaps=11)
```


