---
title: "Class17 Vax Mini Project"
author: 'Monica Lin (PID: A15524235)'
date: "11/25/2021"
output:
  word_document: default
  pdf_document: default
---

# Background

The goal of this hands-on mini-project is to examine and compare the Covid-19 vaccination rates around San Diego.

We will start by downloading the most recently dated “Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code


Move the downloaded CSV file to the Class17 project directory, then read/import into an R object named `vax`. Use this data to answer all the questions below.

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> **Q1**. What column details the total number of people fully vaccinated?

The column "persons_fully_vaccinated" details the total number of people fully vaccinated.

> **Q2**. What column details the Zip code tabulation area?

"zip_code_tabulation_area".

> **Q3**. What is the earliest date in this dataset?

```{r}
head(vax$as_of_date)
```

The earliest date in the dataset is 2021-01-05, by Year-month-date.

> **Q4**. What is the latest date in this dataset?

```{r}
tail(vax$as_of_date)
```

The latest date in this dataset is 2021-11-23.


Let's call the `skim()` function from the **skimr** package to get a quick overview of this dataset.

```{r}
library(skimr)
skimr::skim(vax)
```


> **Q5**. How many numeric columns are in this dataset?

9

> **Q6**. Note that there are "missing values" in the dataset. How many `NA` values are there in the `persons_fully_vaccinated` column?

```{r}
sum( is.na(vax$persons_fully_vaccinated) )
```

There are 8355 NA values in that column.

> **Q7**. What percent of `persons_fully_vaccinated` values are missing (to two significant figures)?

```{r}
sum( is.na(vax$persons_fully_vaccinated) ) / nrow(vax)
```

10.08% of `persons_fully_vaccinated` values are missing.

> **Q8**. [Optional]: Why might this data be missing?

Optional.

# Working with dates

One of the “character” columns of the data is `as_of_date`, which contains dates in the Year-Month-Day format.

Dates and times can be annoying to work with at the best of times. However, in R we have the excellent **lubridate** package, which makes life a lot easier when dealing with dates and times. Here is a quick example to get you started:

```{r}
# install.packages("lubridate")
library(lubridate)
```

What is today's date?

```{r}
today()
```

The `as_of_date` column of our data is currently not that usable. For example, we can't easily do math with it like answering the simple question of how many days have passed since data was first recorded:

However, if we convert our date data into a lubridate format, this like this will be much easier (as well as plotting time series data later on).

```{r}
# Specify that we are using the Year-month-day format
vax$as_of_date <- ymd(vax$as_of_date)
```


Now we can do math with dates. For example: How mnay days have passed since the first vaccination reported in this dataset?

```{r}
today() - vax$as_of_date[1]
```

Using the last and the first date value, we can now determine how many days the dataset span.

```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> **Q9**. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

It has been 4 days since the last entry.

> **Q10**. How many unique dates are in the dataset (i.e. how many different dates are detailed?)

```{r}
length(unique(vax$as_of_date))
```

There are 47 unique dates in the dataset.


# Working with ZIP codes

One of the numeric columns in the dataset (namely `vax$zip_code_tabulation_area`) are actually ZIP codes -- a postal code used by the United States Postal Service (USPS). In R, we can use the **zipcodeR** package to make working with these codes easier. For example, let's install and thn load up this package to find the centroid of the La Jolla 92037 (i.e. UC San Diego) ZIP code area.

```{r}
# install.packages("zipcodeR")
library(zipcodeR)
```

```{r}
# Find centroid of La Jolla 92037 ZIP code area
geocode_zip('92037')
```

Calculate the distance between the centroids of any two ZIP codes in miles, e.g.

```{r}
zip_distance('92037', '92109')
```

More usefully, we can pull census data about ZIP code areas (including median household income, etc.) For example:

```{r}
reverse_zipcode(c('92037', '92109'))
```

We can use this `reverse_zipcode()` to pull census data later on for any or all ZIP code areas we might be interested in.

```{r}
# Pull data for all ZIP codes in the dataset
zipdata <- reverse_zipcode( vax$zip_code_tabulation_area)
```


# Focus on the San Diego area

Let's now focus in on the San Diego County area by restricting ourselves first to `vax$county == "San Diego"` entries. We have two main choices on how to do this: the first using base R, the second using the **dplyr** package:

```{r}
table(vax$county)
```

```{r}
inds <- vax$county=="San Diego"
head(vax[inds, ])
```

Using the **dplyr** package and its **filter()** function, the code would look like this:

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

Using **dplyr** is often more convenient when we are subsetting across multiple criteria -- for example, all San Diego county areas with a population of over 10,000.

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                  age5_plus_population > 10000)
```


> **Q11**. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

There are 107 distinct ZIP codes listed for San Diego County.

> **Q12**. What San Diego County zip code area has the largest 12+ Population in this dataset?

```{r}
which.max(sd$age12_plus_population)
sd$zip_code_tabulation_area[23]
```

The San Diego County ZIP code area of 92057 has the largest 12+ population in this dataset.


Using **dplyr**, select all San Diego *"county"* entries on *"as_of_date"* "2021-11-09" and use this for the following questions.

```{r}
sd.11.09 <- filter(vax, county=="San Diego" & as_of_date=="2021-11-09")
```

> **Q13**. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”?

```{r}
mean(sd.11.09$percent_of_population_fully_vaccinated, na.rm=TRUE)
```

The overall average "Percent of Population Fully Vaccinated" value is 67.34714%.

We can look at the 6-number summary.

```{r}
summary(sd.11.09$percent_of_population_fully_vaccinated)
```


> **Q14**. Using either ggplot or base R graphics, make a summary figure that show the distribution of Percent of Population Fully Vaccinated values as of "2021-11-09"?

Using base R plots:

```{r}
hist(sd.11.09$percent_of_population_fully_vaccinated,
     main="Histogram of Vaccination Rates Across San Diego County",
     xlab="Percent Fully Vaccinated on 2021-11-09",
     ylab="Frequency")
```


Using ggplot:

```{r}
library(ggplot2)

ggplot(sd.11.09) +
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins=10) +
  labs(x="Percent Fully Vaccinated on 2021-11-09", y="Count (ZIP Code Areas",
       title="Histogram of Vaccination Rates Across San Diego County")
```


# Focus on UCSD/La Jolla

UC San Diego resides in the 92037 ZIP code area nd is listed with an age 5+ population size of 36,144.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area == "92037")

ucsd[1,]$age5_plus_population
```

> **Q15**. Using **ggplot**, make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) + aes(as_of_date, percent_of_population_fully_vaccinated) + 
  geom_point() + geom_line(group=1) + ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated", 
       title="Vaccination Rate for La Jolla CA 92037")
```

This plot shows an initial slow roll out in January into Febuary (likely due to limited vaccine availability). This is followed with rapid ramp up until a clear slowing trend from June time onward. Interpertation beyond this requies context from other zip code areas to answer questions such as: is this trend representative of other areas? Are more people fully vaccinated in this area compared to others? Etc.


# Comparing 92037 to other similar sized areas

Let’s return to the full dataset and look across every zip code area with a population at least as large as that of 92037 on *as_of_date* “2021-11-16”.

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                   as_of_date == "2021-11-16")

head(vax.36)
```

> **Q16**. Calculate the mean *“Percent of Population Fully Vaccinated”* for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* “2021-11-16”. Add this as a straight horizontal line to your plot from above with the `geom_hline()` function.

```{r}
vaccination.36 <- mean(vax.36$percent_of_population_fully_vaccinated)
```

```{r}
ggplot(ucsd) + aes(as_of_date, percent_of_population_fully_vaccinated) + 
  geom_point() + geom_line(group=1) + ylim(c(0,1)) + 
  labs(x="Date", y="Percent Vaccinated", 
       title="Vaccination Rate for La Jolla CA 92037") + 
  geom_hline(yintercept=vaccination.36, color="red", linetype="dashed")
```


> **Q17**. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the *“Percent of Population Fully Vaccinated”* values for ZIP code areas with a population as large as 92037 (La Jolla) *as_of_date* “2021-11-16”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```


> **Q18**. Using ggplot, generate a histogram of this data.

```{r}
ggplot(vax.36) + aes(percent_of_population_fully_vaccinated) + 
  geom_histogram(bins=40) + labs(x="Percent Vaccinated", y="Count")
```


> **Q19**. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)

vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

The 92109 ZIP code area is above the average value of 0.6630. However, the 92040 ZIP code area is below the average value.


> **Q20**. Finally, make a time course plot of vaccination progress for all areas in the full dataset with a `age5_plus_population > 36144`.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)


ggplot(vax.36.all) +
  aes(as_of_date,percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color="blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title="Vaccination Rate Across California",
       subtitle="Only areas with a population above 36K are shown.") +
  geom_hline(yintercept=0.66, linetype="dashed")
```

> **Q21**. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

With the detection of the omicron variant, which is more transmittable than the delta variant, and the combination of the lower-than-expected vaccination rates uncovered in this activity, I feel hesitant about meeting for in-person class next week. Traveling by car is safe enough, but traveling by plane for Thanksgiving is slightly concerning to me.



