---
title: 'Class06: R Functions'
author: 'Monica Lin (PID: A15524235)'
date: "10/14/2021"
output: pdf_document
---

## A play with Rmarkdown
This is some plain text. I can make things **bold**. I can also make things *italic*.

```{r}
# This is an R code chunk!
plot(1:10)
```

## R functions

In today's class, we are going to write a function together that grades some students' work.

Questions for today:

>**Q1**. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]


ctrl+alt+i is a keyboard shortcut to call up R code chunk!
ctrl+return runs the command.
```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```


Let's start with student1 and find their average score.
```{r}
mean(student1)
```

But we want to drop the lowest score...
We could try the **min()** function

```{r}
min(student1)
```

The **which.min()** function looks useful:
```{r}
which.min(student1)
```
This gives the position of the lowest numeric score
```{r}
# This would be the lowest score
student1[which.min(student1)]
```
To drop this value, use minus to print out all the other scores
```{r}
student1[ -which.min(student1)]
```
Now use **mean()** to get the average minus the lowest score
```{r}
mean(student1[ -which.min(student1) ])
```
Let's try this with student2
```{r}
student2
mean(student2[ -which.min(student2) ])
```

student2 would get NA as their average grade, just for one missing homework :-(
We need to remove the NA elements of the vector
```{r}
# which.min(student2)
mean(student2[ -which.min(student2)], na.rm=TRUE)
```
This is still not what we want. It dropped the 80 (i.e. the lowest number) instead of the NA (i.e. the missing homework).

Let's look at student3
```{r}
student3
```
na.rm=TRUE would remove all of the NAs and returns NaN (i.e. Not a Number)
```{r}
mean(student3[ -which.min(student3)], na.rm=TRUE)
```
One approach to solve this is to replace any NAs with zeroes.

Let's try with student2
```{r}
student2
is.na(student2)
```
The **is.na()** function returns a logical vector where TRUE elements represent positions of NA values.

which(is.na(x)) gives the position of NA
```{r}
which(is.na(student2))
```

Now let's make the NA values into zeros.
```{r}
student.prime <- student2
student.prime
student.prime[which(is.na(student.prime)) ] = 0
student.prime
```

Now we need to put this all together to get the average score, dropping the lowest where we map NA values to zero.
```{r}
student.prime <- student2
student.prime[ which(is.na(student.prime)) ] = 0
mean(student.prime[ -which.min(student.prime) ])
```
```{r}
student.prime
```
```{r}
mean(c(100,90,90,90,90,97,80))
```

Looks good! Check student3
```{r}
student.prime <- student3
student.prime[ which(is.na(student.prime)) ] = 0
mean(student.prime[ -which.min(student.prime) ])
```

We got our working snippet! Let's simplify.
```{r}
x <- student3
# Map NA values to zero
x[ which(is.na(x)) ] = 0
# Find the mean without the lowest value
mean(x[ -which.min(x) ])
```


Now we can use this as the body of our function.
```{r}
grade <- function(x) {
  # Make sure our scores are all numbers
  x <- as.numeric(x)
  
  # Map NA values to zero
  x[ which(is.na(x)) ] = 0
  # Find the mean without the lowest value
  mean(x[ -which.min(x) ])
}
```
```{r}
grade(student3)
```
The function works!


Now read the full gradebook CSV file.
```{r}
scores <- read.csv("https://tinyurl.com/gradeinput", row.names=1)
scores
```
```{r}
is.numeric(student1)
```
```{r}
is.numeric(scores[10,])
```
```{r}
as.numeric(c(1,2,NA,4,5))
```

Use for one student
```{r}
grade(scores[1,])
```
Change scores to numeric values by altering previous code.
**x <- as.numeric(x)** to avoid having to change each one individually via **grade(as.numeric(scores[2,]))**
Row names must also be set as numeric values, otherwise "student-X" will be read as hw data. **row.names=1**


Now grade all students by using the **apply()** function
```{r}
apply(scores,1,grade)
ans <- apply(scores,1,grade)
```


>**Q2**. Using your grade() function and the supplied gradebook, Who is the top scoring student
overall in the gradebook? [3pts]

```{r}
which.max(ans)
```

>**Q3**. From your analysis of the gradebook, which homework was toughest on students (i.e. obtained
the lowest scores overall? [2pts]

We can use the apply() function over the columns by setting the margin=2 argument. To use the apply() function over the rows, set margin=1.
```{r}
apply(scores,2,mean, na.rm=TRUE)
difficult <- apply(scores,2,mean, na.rm=TRUE)
```
HW3 was toughest.




>**Q4**. Optional Extension: From your analysis of the gradebook, which homework was most
predictive of overall score (i.e. highest correlation with average grade score)? [1pt]

```{r}
mean(ans)
difficult - mean(ans)
```
HW1 was most predictive of overall score because HW1 returns a value closest to 0, meaning that homework average is least different from the class average.


Alternate method to answer **Q4**.
```{r}
mask <- scores
mask[is.na(mask)] = 0
mask

cor(ans, mask$hw3)
```

Do for all with apply
```{r}
apply(mask,2,cor,ans)
```
Using the correlation method, HW5 gives the highest value, so HW5 correlates most strongly and is most predictive of overall score.


# 10/19/2021 Class 07
Can we make a new code chunk so that it does not drop the lowest score?
```{r}
grade2 <- function(x, drop.lowest=TRUE){
  # Make sure our scores are all numbers
  x <- as.numeric(x)
  
  # Map NA values to zero
  x[ - which(is.na(x)) ] = 0
  
  if(drop.lowest) {
    # Find the mean without the lowest value
    mean(x[ - which.min(x) ])
  } else {
    mean(x)
  }
}
```

```{r}
student1 <- c(100, 100, 100, 90)
grade2(student1, drop.lowest=TRUE)
grade2(student1, drop.lowest=FALSE)
```

